SIG Proceedings Paper in LaTeX Format

### 2019-09-28 SAC/submission/ACM_SigConf_SAC2020-2020_1.0/includes/discussion.tex
final tex work

\subsubsection{Shuffling and learning rate with SGD}
In this subsubsection we will use the term SGD for SGD, Momentum and NAG, and advanced optimzers for Adagrad and Adam. The models was able to use a higher learning rates when SGD optimizers where combined with input shuffling in comparison to as when not, as shown in Figures \ref{fig:results_sgd}, \ref{fig:results_mom} and \ref{fig:results_nag}. Therefore, arises the questions why these phenomena happen, especially as it did not happen with advanced optimizers.  \\
First, we suspect our batched version of the Skip-Gram model with negative sampling (SGNS), to be at the origin of this phenomena. In consequence to this batched version, when input shuffling is not used a lot of words will appear multiple times in the same batch, as explained in Section \ref{ssec:shuffling}. Therefore, the gradients will be an average of all the training samples. When the input is shuffled, less words will appear multiple time which will make the gradients more precise.\\
Second, advanced optimizers have a way to counter-attack this issue, namely adaptive learning rates, explained in Section \ref{ssec:results_adagrad}. Adaptive learning rates counter attack the issue in the following way: high frequency words  will appear more often multiple times in a batch, they will also have lower learning rates. Therefore the impact of appearing in the same batch will be reduced. \\  The two above arguments could explain why a higher learning rate combined with SGD achieved better results with input shuffling but not when advanced optimizers are used.

\subsubsection{Large differences with NAG and SGD when using shuffling}
SGD and NAG both have very different values when using shuffling in comparison to unshuffled input, as shown in figures \ref{fig:results_sgd} and \ref{fig:results_nag}. We do not only attribute those results to input shuffling but partially also to a good random initialization guess. Due to a lack of time these experiments were not replicate more than once.

\subsection{Comparison to Gensim}
This subsection will compare our finding extensively to Gensim \cite{gensim}. As explained in Section \ref{ssec:gensim}, Gensim is optimized to have a very high throughput, this made it possible to achieve a lot of computations. Furthermore, Gensim provides access to the loss and the resulting word embeddings, which facilitated the comparison process.
This subsection is structured as follows: first it will describe the used configuration of Gensim, second it will compare Gensim to our implementation of the SGNS with the use of SGD and finally compare our model with Adam to Gensim. A graphic showing the comparison of our models to Gensim can be found in Figure \ref{fig:gensim_vs_adam}.

@@

\subsubsection{Gensim vs. SGD}
As stated earlier, we are not going to compare this work to Gensim in run time. Gensim is heavily optimized and written in cython\footnote{https://rare-technologies.com/word2vec-in-python-part-two-optimizing/}, which is 23x faster than plain Numpy. Since we used PyTorch the difference is not that big, but still remains. As shown in Figure \ref{fig:gensim_vs_adam} the convergence time was not the same between our implementation and Gensim. There are different possible reasons why this could be the case:\\ First, our batched approach could hinder performance in terms of convergence time since our loss function is not exactly the same. We compute the loss for multiple training samples by taking the sum over each score which is individually used as a loss by Gensim.\\ Secondly, a difference to our implementation is the fact that Gensim checks whether negative samples are not equal to the context word. If that is the case Gensim selects a new random sample. Therefore, the learning of the input and output context is optimized. \\Finally, another possibility is the decay of the learning rate used by Gensim. In fact, decaying the learning rate has been proven in a lot of work to decrease the convergence time. Gensim linearly decreases the learning rate, as we did not use this technique, the decay of the learning rate could help explain the noted differences. \\ The first hypothesis, the fact that we used a batched approach, may be confirmed by the fact that when combined with input shuffling SGD does perform closer to Gensim, going from 11 to 7 epochs to converge, as input shuffling reduces the number of co-occurrence of the same word in a batch.
Know the question arises if the 3 epochs, that Gensim is better, can be explained by the selection of better negative samples and the learning rate decay.

>>>>

\subsubsection{Shuffling and learning rate with SGD}
In this subsubsection we will use the term advanced optimzers for Adagrad and Adam. The models was able to use a higher learning rates when SGD was combined with input shuffling in comparison to as when not. Therefore, arises the questions why these phenomena happen, especially as it did not happen with advanced optimizers.  \\
First, we suspect our batched version of the Skip-Gram model with negative sampling (SGNS), to be at the origin of this phenomena. In consequence to this batched version, when input shuffling is not used a lot of words will appear multiple times in the same batch, as explained in Section \ref{ssec:shuffling}. Therefore, the gradients will be an average of all the training samples. When the input is shuffled, less words will appear multiple time which will make the gradients more precise.\\
Second, advanced optimizers have a way to counter-attack this issue, namely adaptive learning rates, explained in Section \ref{ssec:results_adagrad}. Adaptive learning rates counter attack the issue in the following way: high frequency words  will appear more often multiple times in a batch, they will also have lower learning rates. Therefore the impact of appearing in the same batch will be reduced. \\  The two above arguments could explain why a higher learning rate combined with SGD achieved better results with input shuffling but not when advanced optimizers are used.

\iffalse
\subsubsection{Large differences with NAG and SGD when using shuffling}
SGD and NAG both have very different values when using shuffling in comparison to unshuffled input, as shown in figures \ref{fig:results_sgd} and \ref{fig:results_nag}. We do not only attribute those results to input shuffling but partially also to a good random initialization guess. Due to a lack of time these experiments were not replicate more than once.
\fi

\subsection{Comparison to Gensim}
This subsection will compare our finding extensively to Gensim \citep{gensim}. As explained in Section \ref{ssec:gensim}, Gensim is optimized to have a very high throughput, this made it possible to achieve a lot of computations. Furthermore, Gensim provides access to the loss and the resulting word embeddings, which facilitated the comparison process.
This subsection is structured as follows: first it will describe the used configuration of Gensim, second it will compare Gensim to our implementation of the SGNS with the use of SGD and finally compare our model with Adam to Gensim. A graphic showing the comparison of our models to Gensim can be found in Figure \ref{fig:gensim_vs_adam}.

@@

\subsubsection{Gensim vs. SGD}
As stated earlier, we are not going to compare this work to Gensim in run time. Gensim is heavily optimized and written in cython\footnote{https://rare-technologies.com/word2vec-in-python-part-two-optimizing/}, which is 23x faster than plain Numpy. Since we used PyTorch the difference is not that big, but still remains. As shown in Figure \ref{fig:gensim_vs_adam} the convergence time was not the same between our implementation and Gensim. There are different possible reasons why this could be the case:\\ First, our batched approach could hinder performance in terms of convergence time since our loss function is not exactly the same. We compute the loss for multiple training samples by taking the sum over each score which is individually used as a loss by Gensim.\\ Finally, another possibility is the decay of the learning rate used by Gensim. In fact, decaying the learning rate has been proven in a lot of work to decrease the convergence time. Gensim linearly decreases the learning rate, as we did not use this technique, the decay of the learning rate could help explain the noted differences. \\ The first hypothesis, the fact that we used a batched approach, may be confirmed by the fact that when combined with input shuffling SGD does perform closer to Gensim, going from 7 to 5 epochs to converge, as input shuffling reduces the number of co-occurrence of the same word in a batch.
Know the question arises if the 3 epochs, that Gensim is better can be explained by the batched approach and the learning rate decay.

### 2019-09-28 SAC/submission/ACM_SigConf_SAC2020-2020_1.0/includes/results.tex
final tex work

\section{Results}\label{sec:results}
As stated in the previous section we ran multiple experiments for each optimizer with the above configuration and datasets. . This Section will  give an overview of the experimental results. Each subsection will give an explanation over the achieved result with a specific optimizer.

\subsubsection{SGD}
The first challenge for each optimizer was to find a correct learning rate. As SGD is the optimizer used in Gensim \citep{gensim} we first tried the same learning rate as the default value in Gensim \citep{gensim}, i.e 0.01,  and then performed a random search to find a better one. As expected a bell curve shape resulted, a learning rate that is too high leads to diversion and a learning rate that is too low leads to a training time that is too slow. The best value that we found for the learning rate is $0.0075$. With this setting SGD converged in 11 epochs. The second experiment was to add input shuffling.
For almost every learning rate the convergence time decreased. Our model, with the best setting, know converges in only 7 epochs. Another interesting fact to point out is that with input shuffling we achieved better results with higher learning rates. As for learning rates of $0.01$ and $0.025$ we did converge in 11 epochs with input shuffling but did not converge in 20 epochs without it.
\iffalse
\begin{figure}[h]
\centering
\includegraphics[scale=0.3]{images/results_sgd_shuffle}
\caption{Training time Stochastic Gradient Descent with input Shuffling}
\label{fig:results_sgd}
\end{figure}
\fi
\iffalse 
\subsubsection{Momentum and Nesterov}
Momentum and  Nesterov Accelerated Gradient (NAG) \citep{nag} both have an additional hyperparameter $\gamma$, that defines the percentage of the previous gradient that will be added to the current gradients. We set $\gamma = 0.9$ as this is a typical value and did not alter it during our experiments. Momentum and NAG alone respectively only slightly decrease or increase the convergence time. Momentum optimally converges in 9 epochs and NAG in 13. If we combine these optimizers with input shuffling, interestingly the same phenomena as with plain SGD appear. The convergence time gets better, 8 epochs for Momentum and 3 epochs for NAG. The phenomena that a higher learning rat yields better results also happens with both of the optimizers. As Momentum does not converge in 20 epochs with a learning rate of 0.002 but does in 8 with input shuffling.
\begin{figure}[h]
\centering

@@

\subsubsection{Adagrad}\label{ssec:results_adagrad}
Adagrad \citep{adagrad} is a very interesting optimizer for creating word embeddings as it decreases the learning rate for very frequent occurring features, and vice versa. Words (features in our case) that appear very frequently (low learning rate) often do not have a semantic gain  that is as important as words that appear less frequently (high learning rate) to their context words. So, in theory, Adagrad is particularly well suited for our task, as for example Pennington et al. used Adagrad in the training of Glove \citep{glove}, another system which word embeddings. Our results confirmed empirically that Adagrad can work better for creating word embeddings. The  model converged in 4 epochs. When combined with shuffling Adagrad only took 3 epochs to converge, as shown in Table \ref{table:results}. These results confirms two hypothesis: first  the tendency of the Skip-Gram Model to converge faster with input shuffling and second  the big impact of having different learning rate for each feature.
Another interesting fact to notice is that a higher learning rate combined with input shuffling did not yield better results than without shuffling, in comparison to the earlier described optimzers. Adagrad performed best with a learning rate of $0.1$.
\iffalse
\begin{figure}[h]
\centering
\includegraphics[scale=0.3]{images/results_adagrad_shuffle}
\caption{Training time Adagrad with input Shuffling}
\label{fig:results_adagrad_shuffle}
\end{figure}
\fi
\subsubsection{Adam}
Adam is the most advanced of all the optimizers used in our experiments and did yield the best results as seen in F. Adam converged in 3 epochs without shuffling and a learning rate of $0.05$ and 2 with shuffling and a learning rate of $0.001$, as shown in Table \ref{table:results}. Those are the best results achieved in any of our experiment. Exacctly as Adagrad, Adamd did not react to input shuffling the same way as SGD did.
    \begin{table}[]
\begin{tabular}{|l|l|l|}
\hline
Model    & Convergence Time & Word Similarity \\ \hline
SGD & {11}              & 0.65            \\ \hline
SGD w/shuffling & {7}              & 0.66            \\ \hline
Adam & {3}              & 0.66            \\ \hline
Adam w/ shuffling & \textbf{2}              & 0.66      \\ \hline
\end{tabular}
\label{table:results}
\caption{Results on the text8 dataset with different optimizers and input shuffling}

>>>>

\section{Results}\label{sec:results}
As stated in the previous section we ran multiple experiments for each optimizer with the above configuration and datasets. This Section will  give an overview of the experimental results. Each subsection will give an explanation over the achieved result with a specific optimizer.

\subsubsection{SGD}
The first challenge for each optimizer was to find a correct learning rate. As SGD is the optimizer used in Gensim \citep{gensim} we first tried the same learning rate as the default value in Gensim \citep{gensim}, i.e 0.025,  and then performed a random search to find a better one. As expected a bell curve shape resulted, a learning rate that is too high leads to diversion and a learning rate that is too low leads to a training time that is too slow. The best value that we found for the learning rate is $0.0075$. With this setting SGD converged in 7 epochs. The second experiment was to add input shuffling.
For almost every learning rate the convergence time decreased. Our model, with the best setting, know converges in only 5 epochs. 
\iffalse
\begin{figure}[h]
\centering
\includegraphics[scale=0.3]{images/results_sgd_shuffle}
\caption{Training time Stochastic Gradient Descent with input Shuffling}
\label{fig:results_sgd}
\end{figure}
\fi
\iffalse 
\subsubsection{Momentum and Nesterov}
Momentum and  Nesterov Accelerated Gradient (NAG) \citep{nag} both have an additional hyperparameter $\gamma$, that defines the percentage of the previous gradient that will be added to the current gradients. We set $\gamma = 0.9$ as this is a typical value and did not alter it during our experiments. Momentum and NAG alone respectively only slightly decrease or increase the convergence time. Momentum optimally converges in 9 epochs and NAG in 13. If we combine these optimizers with input shuffling, interestingly the same phenomena as with plain SGD appear. The convergence time gets better, 8 epochs for Momentum and 3 epochs for NAG. The phenomena that a higher learning rat yields better results also happens with both of the optimizers. As Momentum does not converge in 20 epochs with a learning rate of 0.002 but does in 8 with input shuffling.
\begin{figure}[h]
\centering

@@

\subsubsection{Adagrad}\label{ssec:results_adagrad}
Adagrad \citep{adagrad} is a very interesting optimizer for creating word embeddings as it decreases the learning rate for very frequent occurring features, and vice versa. Words (features in our case) that appear very frequently (low learning rate) often do not have a semantic gain  that is as important as words that appear less frequently (high learning rate) to their context words. So, in theory, Adagrad is particularly well suited for our task, as for example Pennington et al. used Adagrad in the training of Glove \citep{glove}, another system which word embeddings. Our results confirmed empirically that Adagrad can work better for creating word embeddings. The  model converged in 4 epochs. When combined with shuffling Adagrad only took 3 epochs to converge, as shown in Table \ref{table:results}. These results confirms two hypothesis: first  the tendency of the Skip-Gram Model to converge faster with input shuffling and second  the big impact of having different learning rate for each feature.
Another interesting fact to notice is that a higher learning rate combined with input shuffling did not yield better results than without shuffling, in comparison to the earlier described optimzers. Adagrad performed best with a learning rate of $0.1$.
\iffalse
\begin{figure}[h]
\centering
\includegraphics[scale=0.3]{images/results_adagrad_shuffle}
\caption{Training time Adagrad with input Shuffling}
\label{fig:results_adagrad_shuffle}
\end{figure}
\fi
\subsubsection{Adam}
Adam is the most advanced of all the optimizers used in our experiments and did yield the best results as seen in Table \ref{table:results}. Adam converged in 3 epochs without shuffling and a learning rate of $0.05$ and 2 with shuffling and a learning rate of $0.001$, as shown in Table \ref{table:results}. Those are the best results achieved in any of our experiment. Exacctly as Adagrad, Adamd did not react to input shuffling the same way as SGD did.
    \begin{table}[]
\begin{tabular}{|l|l|l|}
\hline
Model    & Convergence Time & Word Similarity \\ \hline
SGD & {11}              & 0.65            \\ \hline
SGD w/shuffling & {7}              & 0.66            \\ \hline
Adam & {3}              & 0.66            \\ \hline
Adam w/ shuffling & \textbf{2}              & 0.66      \\ \hline
\end{tabular}
\label{table:results}
\caption{Results on the text8 dataset with different optimizers and input shuffling}
